---
title: "Report"
author: "Vasco Ferreira"
date: "7/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This data analytics report is one the Capstone Exercises for the Google Data
Analytics Course on Coursera. This case study is about a bike-sharing company
that aims to increase the number of annual members (subscribers) as these type 
of members are the most profitable for the company.

Information about the company:

- 5800 bikes
- 600 docking stations
- different types of bikes:
  - traditional bikes
  - reclining bikes
  - hand tricycles
  - cargo bikes
- 92% of members use traditional bikes
- 8% of members use the other types of bikes
- 70% of members use bikes for leisure
- 30% of members use bikes for commuting

## Load libraries and data

```{r}
library(tidyverse)
```

```{r}
tripdata <- read_csv("../data/processed/02_tripdata.csv")
```


## Ask

The task is to answer the first question of the new marketing program:

> How do annual members and casual riders use Cyclistic bikes differently?

```{r}
tripdata %>% 
  drop_na(usertype) %>% 
  group_by(usertype) %>% 
  summarize(average_trip_duration = mean(calculated_tripduration), 
            std_deviation = sd(calculated_tripduration),
            min_trip_duration = min(calculated_tripduration),
            max_trip_duration = max(calculated_tripduration))
```

In general, the