---
title: "Report"
author: "Vasco Ferreira"
date: "7/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This data analytics report is one the Capstone Exercises for the Google Data
Analytics Course on Coursera. This case study is about a bike-sharing company
that aims to increase the number of annual members (subscribers) as these type 
of members are the most profitable for the company.

Information about the company:

- 5800 bikes
- 600 docking stations
- different types of bikes:
  - traditional bikes
  - reclining bikes
  - hand tricycles
  - cargo bikes
- 92% of members use traditional bikes
- 8% of members use the other types of bikes
- 70% of members use bikes for leisure
- 30% of members use bikes for commuting

## Load libraries and data

```{r}
library(tidyverse)
library(lubridate)
library(tidyquant) 
```

Define the types for the data as the parser only looks for the first rows
and there is data with a new format in the later rows:

```{r}
col_types = list(
  col_character(),
  col_character(),
  col_datetime(),
  col_datetime(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_double(),
  col_double(),
  col_double(),
  col_double(),
  col_character(),
  col_double(),
  col_double()
)
```

Load the data:

```{r}
tripdata <- read_csv("../data/processed/02_tripdata.csv", col_types = col_types)
```

The task is to answer the first question of the new marketing program:

> How do annual members and casual riders use Cyclistic bikes differently?

## How many users have a membership status?

To answer this question we need to first group the data by each member type
and then calculate some basic descriptive statistics:

```{r}
tripdata %>% 
  drop_na(member_casual) %>%  
  ggplot() +
    geom_bar(aes(member_casual, y=stat(count/sum(count)), fill=member_casual)) +
    ylab("Percentage of users") +
    xlab("Type of user") +
    labs(title="How many of our users have a membership status?")
```

In this graph we can see that almost 60% of our users have a membership
status.

## What is the most used type of bike?

```{r}
tripdata %>% 
  drop_na(member_casual, rideable_type) %>%  
  ggplot() +
    geom_bar(aes(rideable_type, y=stat(count/sum(count)), fill=rideable_type)) +
    facet_wrap(~member_casual) +
    ylab("Percentage of users") +
    xlab("Type of bike") +
    labs(title="What type of bike do our users use the most")
```

Both types of users use the `docked bike` the most but, the `member` users
use the `classic bike` more than the `electrict bike`. The `casual` users prefer
the `electric bike` over the `classic bike`

## For how long do our users travel?

```{r}
tripdata %>% 
  drop_na(member_casual, trip_duration) %>% 
  filter(trip_duration > 0, trip_duration < 7500) %>% 
  ggplot() +
    geom_histogram(aes(x=trip_duration, y = stat(count / sum(count)), fill=member_casual)) +
    facet_wrap(~member_casual) +
    ylab("Percentage of users") +
    xlab("Trip duration in seconds") +
    labs(title="Distribution of trip duration by user type")
```

```{r}
tripdata %>% 
  drop_na(member_casual, trip_duration) %>%
  ggplot() +
    geom_boxplot(aes(trip_duration, fill=member_casual)) +
    coord_cartesian(xlim=c(0, 2500)) +
    ylab("Percentage of users") +
    xlab("Trip duration in seconds") +
    labs(title="Distribution of trip duration by user type")
```

The average `casual` user spends double of the
time of a `member` during a trip, 20 minutes versus 11 minutes. And that
the trip duration is more spread out among `casual` members. This could be a
demonstration of the majority of the users using bike for leisure trips as
indicated in basic information.

## How far do our users travel?

Although the distance calculated through the latitude and longitude isn't the
most accurate representation of the distance traveled as trips may not be in
a straight line, it is the only representation that we currently have.

```{r}
tripdata %>% 
  drop_na(member_casual, distance_traveled) %>% 
  filter(distance_traveled > 0, distance_traveled < 10000) %>% 
  ggplot() +
    geom_histogram(aes(x=distance_traveled, y = stat(count / sum(count)), fill=member_casual)) +
    facet_wrap(~member_casual) +
    ylab("Percentage of users") +
    xlab("Distance traveled in meters") +
    labs(title="Distribution of distance traveled by user type")
```

```{r}
tripdata %>% 
  drop_na(member_casual, distance_traveled) %>%  
  ggplot() +
    geom_boxplot(aes(distance_traveled, fill=member_casual)) +
    coord_cartesian(xlim=c(0, 2500)) +
    ylab("Percentage of users") +
    xlab("Distance traveled in meters") +
    labs(title="Distribution of distance traveled by user type")
```

The average `casual` and `member` users
travel about the same distance, however, the `member` users' shorter trips are longer
than the `casual` users'.

## When do our users travel?

```{r}
tripdata %>% 
  mutate(weekday = weekdays(started_at, abbreviate = TRUE)) %>%
  mutate(weekday_sorted = factor(weekday, levels=c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))) %>% 
  drop_na(member_casual, weekday_sorted) %>%
  ggplot() +
    geom_bar(aes(weekday_sorted, y = stat(count / sum(count)), fill=member_casual)) +
    facet_wrap(~member_casual) +
    xlab("Weekday") +
    ylab("Percentage of users") +
    labs(title="Users per day of week")
```

In this graph we notice that `member` users most probably use our bikes
for commuting during the week as oposed to `casual` members who use our bikes
more at weekends.

## How many users do we have through the year?

```{r}
tripdata %>% 
  mutate(year = year(started_at),
         month = month(started_at),
         day = day(started_at),
         date = date(started_at)) %>%
  drop_na(member_casual, date) %>% 
  group_by(date, member_casual) %>%
  arrange(date) %>%
  summarize(number_users = n()) %>%
  ggplot() +
    geom_ma(aes(date, number_users, color=member_casual), n=14, linetype="solid") +
    xlab("Date") +
    ylab("Number of users") +
    labs(title="Number of users over the year (14 days moving average)")
```

```{r}
tripdata %>% 
  mutate(year = year(started_at),
         month = month(started_at),
         day = day(started_at),
         date = date(started_at)) %>%
  drop_na(member_casual, date) %>% 
  group_by(date, member_casual) %>%
  arrange(date) %>%
  summarize(average_duration = mean(distance_traveled)) %>%
  ggplot() +
    geom_ma(aes(date, average_duration, color=member_casual), n=14, linetype="solid") +
    xlab("Date") +
    ylab("Distance in meters") +
    labs(title="Average distance traveled over the year (14 days moving average)")
```


It is very difficult to draw conclusions from this graph for the simple fact 
that this is not a normal year as the world was affected by COVID-19 and 
COVID-19 also has trends. However, we can clearly see a trend in users over the 
year so that in the summer there are more users than in the rest of year.

## Conclusion






