---
title: "02_data_cleaning"
author: "Vasco Ferreira"
date: "7/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries needed

```{r}
library(tidyverse)

# Used to apply functions in parallel to rows
library(purrr)

# Used to calculate distances with latitude and longitude
library(geosphere)
```

## Load the data

```{r}
# Use the column types to avoid parsing errors
col_types = list(
  col_character(),
  col_character(),
  col_datetime(),
  col_datetime(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_double(),
  col_double(),
  col_double(),
  col_double(),
  col_character()
)

tripdata <- read_csv("../data/processed/01_tripdata.csv", col_types = col_types)
```

## Check the data

```{r}
summary(tripdata)
```

## Clean the data

### Trip duration

Calculate the trip duration to check if there are `started_at` and `ended_at`
dates that are swapped and check the summary of the data:

```{r}
cleaned_tripdata <- tripdata %>% 
  mutate(trip_duration = as.numeric(difftime(ended_at, 
                                             started_at, 
                                             units = "secs")))

summary(cleaned_tripdata)
```

By checking the the summary of the data, we can see that at least the 
`trip_duration` probably has bad data as the maximum value is `3523202`, which
is much greater than the average value and that there are negative values that 
implies that either the time is going backwards or swapped values between the 
`started_at` and the `ended_at`.

```{r}
fixed_rows <- cleaned_tripdata %>% 
  filter(trip_duration <= 0) %>% 
  arrange(trip_duration) %>% 
  mutate(temp = started_at) %>% 
  mutate(started_at = ended_at, ended_at = temp) %>% 
  select(-temp)

cleaned_tripdata <- cleaned_tripdata %>% 
  filter(trip_duration > 0) %>% 
  bind_rows(fixed_rows) %>% 
  mutate(trip_duration = as.numeric(difftime(ended_at, 
                                             started_at, 
                                             units = "secs")))
```

Check to see if the negative numbers are fixed:

```{r}
summary(cleaned_tripdata)
```

Let's check why the maximum value is so high:

```{r}
cleaned_tripdata %>% 
  filter(trip_duration > 1000000)
```

By checking the data, the maximum value of the `trip_duration` is
so great because the client kept the bike for more than a month. This means
that there isn't bad data.

### User type

By checking the `usertype` column we can also see that there doesn't seem
to exist any bad data.

```{r}
cleaned_tripdata %>% 
  ggplot() +
    geom_bar(aes(member_casual))
```

## Stations

Check if there is missing or bad data in the station names

```{r}
cleaned_tripdata %>% 
  group_by(start_station_id) %>% 
  summarize(number_trips = n(),
            average_duration = mean(trip_duration),
            deviation = sd(trip_duration),
            min_duration = min(trip_duration),
            max_duration = max(trip_duration)) %>% 
  arrange(-number_trips)
```

Let's keep the data that have at least the latitude and longitude so we
can calculate the distance traveled by customers.

```{r}
cleaned_tripdata <- cleaned_tripdata %>%  
  filter(!is.na(start_lat) & !is.na(end_lat))
```

### Distance traveled

Calculate the distance traveled as it can be useful in the analysis (**this code
takes a while to run**)

```{r}
calculate_distance <- function(lon1, lat1, lon2, lat2) {
  dist <- distm(c(lon1, lat1), c(lon2, lat2), fun = distHaversine)
  return(dist)
}

cleaned_tripdata <- cleaned_tripdata %>% 
  mutate(distance_traveled = pmap_dbl(list(start_lat, start_lng, end_lat, end_lng), calculate_distance))
```

Check for negative distances

```{r}
cleaned_tripdata %>% 
  filter(distance_traveled < 0)
```

## Save the cleaned data

Check the columns to be saved:

```{r}
colnames(cleaned_tripdata)
```

```{r}
str(cleaned_tripdata)
```


As we checked the data and cleaned what seemed to be bad data, we now save
the data and use it in the next steps of the data analysis.

```{r}
write_csv(cleaned_tripdata, "../data/processed/02_tripdata.csv")
```

